A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-11-15 thesis/thesis.tex
finalize thesis

\section*{Acknowledgements}
\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed. He helped me in every step of this thesis, taught me the shared data structures, helped me find the problem, shape up the algorithm, and write it with its complications because of lots of new notions in it. Thank you again for always pushing to make it better.

@@

Last but not the least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study. I want to thank my brother for being there whenever I needed help.

@@

I dedicate this paper to all the people who are fighting for their freedom.

@@

\pagenumbering{arabic}
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same. In this run, $P$'s update is overwritten, which makes queue inconsistent. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty, it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for the implementation of a shared data structure?''. Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

@@

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also discuss the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself and there is an example of a concurrent exuection on the algorithm described in Section 3.3.

@@

Chapter 5 is devoted to complexity analysis for our implementation. We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end, we prove our queue is wait-free.

@@

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. The worst-case step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). A shared variable Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption[MS-queue structure]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram, the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes' enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations. However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the worst-case step complexity remains $\Omega(p)$ per operation.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption[Baskets queue]{At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}. To order the operations in a basket, processes perform another \nf{CAS}. The successful process will be the next one in the basket, and so on.}
\end{figure}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one. The worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is full, an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}). A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}]{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their worst-case step complexity is $\Omega(p)$ because of the helping mechanism.

@@

Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. The fast path  ensures good performance while the slow path ensures termination. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

The CAS Retry Problem is not limited to list-based queues; array-based queues also share it as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  Our motivation is to overcome this problem and present a wait-free sublinear queue.

@@

\subsection{Universal Constructions and Other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. However their algorithm cannot be used to create a queue. We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of the Page 13.

@@

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes. When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree. There are some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues. Also, it is one of the few wait-free data structures achieving poly-log complexity.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that since their result is about when contention is low, their lower bound does not contradict with our algorithm's complexity and we manage to reach poly-log time complexity.

@@

There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes. The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree})Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption[Tournament tree]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

@@

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{After the \nf{Refresh}}
\end{subfigure}
\hspace{6em}
\caption[A successful \nf{Refresh}]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
  \caption[Two consecutive failed \nf{Refresh}es by a process]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
\end{figure}

@@

\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]
\end{tikzpicture}
\caption[Showing the operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right. In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together. Since their execution time intervals overlap, they can be linearized in any order.}
\end{center}
\end{figure}

@@

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption[Showing the operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

@@

\begin{algorithm}
\caption{\tt{\sl{Root}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{36}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within the \nf{b}th block in the \nf{root}.}

>>>>

\section*{Acknowledgements}
\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed. He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it. Thank you again for always pushing to make it better.

@@

Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study. I want to thank my brother for being there for me whenever I needed help.

@@

I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.

@@

\pagenumbering{arabic}
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same. In this run, $P$'s update is overwritten, which makes the queue inconsistent. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty, it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

@@

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem are presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also discuss the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.

@@

Chapter 5 is devoted to a complexity analysis for our implementation. We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end, we prove our queue is wait-free.

@@

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram, the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes' enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations. However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}. To order the operations in a basket, processes perform another \nf{CAS}. The successful process will be the next one in the basket, and so on.}
\end{figure}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one. The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is full, an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}). A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.

@@

Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. The fast path  ensures good performance while the slow path ensures termination. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the amortized complexity cannot be better than $\Omega(p)$.

The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  Our motivation is to overcome this problem and present a wait-free sublinear queue.

@@

\subsection{Universal Constructions and Other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. However, their algorithm cannot be used to create a queue. We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.

@@

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes. When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree. There are some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues. Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.

@@

There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes. The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

@@

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{After the \nf{Refresh}.}
\end{subfigure}
\hspace{6em}
\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
\end{figure}

@@

\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]
\end{tikzpicture}
\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right. In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together. Since their execution time intervals overlap, they can be linearized in any order.}
\end{center}
\end{figure}

@@

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

@@

\begin{algorithm}
\caption{\tt{\sl{Root}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{36}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
